---
title: "Does Technology Advancement Decrease Distracted Driving?" 
author: 
  - Harrison Huang
thanks: "Code and data are available at: https://github.com/lemonface88/Distracted-Driving-Analysis"
date: today
date-format: long
abstract: "Distracted driving is often a leading factor of traffic accidents. Using data from OpenDataToronto, this paper will look at how the improvements of technology over the years from 2016 contribute to the number of distracted driving violations. This analysis discovered with technology improving year by year, the overall trend of distracted driving is showing a decreasing trend."

format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(knitr)
library(tinytex)
library(opendatatoronto)
library(dplyr)
library(janitor)
library(kableExtra)
```


```{r}
#| include: false
#| echo: false
#| message: false

# read cleaned youth data
cleaned_youth_data = read_csv(
  file = here("outputs/data/cleaned_youth_data.csv"),
  show_col_types = FALSE
)

# read cleaned adult data
cleaned_adult_data = read_csv(
  file = here("outputs/data/cleaned_adult_data.csv"),
  show_col_types = FALSE
)

# read cleaned ticket data
 cleaned_ticket_data = read_csv(
  file = here("outputs/data/cleaned_ticket_data.csv"),
  show_col_types = FALSE
)

# read adult and youth data 
sum_total_data = read_csv(
  file = here("outputs/data/sum_total_data.csv"),
  show_col_types = FALSE
)
```

# Introduction
As technology is improving everyday year by year, does it help us reduce distracted driving cases or does it increase the distractions in our surroundings? Distracted driving has been the root of many traffic accidents and caused many unfortunate events since personal handheld devices were introduced. With the advancement of technology, how has this been improved or has it worsened? To examine the impact of distracted driving in our everyday life, I will use the ticket data from open toronto specifically looking at distracted driving tickets, and infer the number of distracted driving occurrences from the ticket data. 

# Data

Data used in this paper are retrieved from Open Data Toronto Portal through the library `opendatatoronto` [@R-opendatatoronto]. Data was cleaned and analyzed using the open source statistically programming language R [@R-base], using functionalities from `tidyverse` [@R-tidyverse], `ggplot2` [@R-ggplot2], `dplyr` [@R-dplyr], `readr` [@R-readr], `tibble` [@R-tibble], `janitor` [@R-janitor],  and `knitr` [@R-knitr]. Details of the data extraction and cleaning processes are discussed in the subsections below.


## Below is a table of the cleaned and filtered table of all distracted driving data
The table is a representation of the selected columns that remain in the data set after cleaning. This table gives insight to the first 5 rows that are left in the distracted driving tickets handed out. 
```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_ticket_statistics
#| tbl-cap: Sample of Cleaned ticket Statistics Data

head (cleaned_ticket_data, 5) |>
  kable(
    col.names = c("Year", "Offence Type", "Age Group", "Ticket Count"),
  booktabs = TRUE
  )
```

## Violations by age groups

So what about distracted driving violations? To dive deeper into to the data analysis, the table is then separated into a table that only includes either all youth violations or adult violations. @tbl-cleaned_youth_statistics is the version that only includes youth distracted driving data, and @tbl-cleaned_adult_statistics is the other version that only includes distracted driving.


```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_youth_statistics
#| tbl-cap: Sample of Cleaned youth distracted driving

only_youth_data <-
  cleaned_youth_data |>
  arrange(offence_year)|>
  select(offence_year, ticket_count)
  
# each individual ticket count represents a different neighborhood
head(only_youth_data, 5) |>
  kable(
    col.names = c("Year", "Ticket Count"),
  booktabs = TRUE
  )
```


```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_adult_statistics
#| tbl-cap: Sample of Cleaned adult distracted driving

only_adult_data <-
  cleaned_adult_data |>
  arrange(offence_year)|>
  select(offence_year, ticket_count)

head (only_adult_data, 5) |>
  kable(
    col.names = c("Year", "Ticket Count"),
  booktabs = TRUE
  )
```

  
# Results

## Violations by youth group per year

The graph below shows the total distracted driving violations cases in each year, which is inferred from the total number of tickets given for distracted driving. Overall, the trend sees a big dip in the year 2020 and 2021, this could be due to the effect of covid. As we all know, the lock down that happened during the pandemic prevented people from leaving the house, thus might have caused the overall decrease in the number of violations in those years. 

```{r}
  only_youth_data |> 
  summarize(sum_per_year = sum(ticket_count, na.rm = TRUE),.by = offence_year) |>
  
  ggplot(mapping = aes(x = offence_year, y = sum_per_year))+ 
  geom_bar(stat= 'identity') +
  scale_x_continuous(breaks =c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022))+ 
  labs(title = "Youth Distracted Driving violations Per year", x = "Year", y = "total ticket")
```

## Violations by adult group per year
  The graph below shows the total distracted driving violations cases committed by adults in each year, which is inferred from the total number of tickets given for distracted driving. Overall, the trend sees a big dip in the year 2020 and 2021, this displays a similar trend as the youth drivers. With the two groups of driving having overall lower violations in the year 2020 and 2021, it could suggest that those years might have some other contributing factors involved other than the advancement of technology. 

```{r}
  only_adult_data |> 
  summarize(sum_per_year = sum(ticket_count, na.rm = TRUE),.by = offence_year) |>
  
  ggplot(mapping = aes(x = offence_year, y = sum_per_year))+ 
  geom_bar(stat= 'identity') +
  scale_x_continuous(breaks =c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022))+ 
  labs(title = "Adult Distracted Driving violations Per year", x = "Year", y = "total ticket")
```

## Overall violation by both groups per year

Upon looking at the summary, the graph below shows the total distracted driving violations cases committed by both adults and youth in each year, which is inferred from the total number of tickets given for distracted driving. As absorbed above, 2020 and 2021 both indicate a big decrease in violation numbers compared to all the other years. However, by combining both of the you and adult data, the overall trend observed is that distracted driving violations have steadily decreased during 2015 to 2019. Despite the numbers increasing again in 2022, the total violations compared to 2015 is still less by about 4000 cases.

```{r}
  sum_total_data |> 
  ggplot(mapping = aes(x = offence_year, y = total_violations))+ 
  geom_bar(stat= 'identity') +
  scale_x_continuous(breaks =c(2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023))+ 
  labs(title = "Both Groups Distracted Driving violations Per year", x = "Year", y = "total ticket")
```

# Discussion
  Simply looking at the numbers of distracted driving tickets, the data shows an overall trend of decreasing occurrences. However, the youth group and adult groups show a slightly different trend with the youth group not showing a clear decreasing trend. A worthy mention is that the amount of youth data could be contributing to the different trend, because there is very little youth data compared to the adult data. In the future, it would be worthwhile to look at other data sets such as integrated technology trends such as how many models of cars have car play or support handless controls. 
# Conclusion
  Overall, the paper examines the number of distracted driving occurrences from 2014 to 2022. Adults show a steady decrease in occurrence with a jump back up in 2022, and the youth data on the contrary shows a steady increase of distracted driving cases. Both data show a dip in 2020 and 2021 with a jump back up in 2022. When looking at the data with both youth group and adult group, we do see an overall decreasing trend of distracted driving cases. With the big jump in 2022, tech companies should look into how to integrate driving and technology better to continue the decreasing trend in the future. 

\newpage

# References